---
title: "EVD2"
author: "Louis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Set-up

```{r data-import, include=FALSE}
library(dplyr)
library(MASS)
library(tidyverse)
library(DescTools)


# Keeps 'random' consistent
set.seed(123)

# Loads SM Volume data
dat <- read.csv('C:\\Users\\Admin\\Documents\\NGED Work\\CDP Smart Meter Data\\Data\\sm-vols-jan24-dummy.csv')
```

```{r filter-outliers}
remove_outliers <- function(df, column) {
  Q1 <- quantile(column, 0.25)
  Q3 <- quantile(column, 0.75)
  IQR <- Q3 - Q1
  
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q1 + 1.5 * IQR
  
  filtered_df <- df[column >= lower_bound & column <= upper_bound,]
  
  return(filtered_df)
}
```

## Turn data into contingency table

```{r contingency-table}
LAs = unique(dat$Licence_Area)


tab_dat = c()
for (LA in LAs) {
  filtered <- dat[dat$Licence_Area == LA,]
  
  S1 = sum(filtered$SMETS1)
  S2 = sum(filtered$SMETS2)
  nonS = sum(filtered$NON_SMART)
  
  row_dat = c(S1, S2, nonS)
  
  tab_dat = append(tab_dat, row_dat)
}

tab <- matrix(tab_dat, nrow=3, ncol=length(LAs), dimnames=list(c("SMETS1","SMETS2","NON SMART"),LAs))

print(tab)
```

Using this table we can test for the independence of the variables 'Licence Area' and 'Meter Type'.

$H_0:$ The meter type is independent of which licence area it resides in.
$vs.$
$H_1:$ The meter type depends on which licence area the meter resides in.

Under $H_0$ for large $n$ (which is applicable here as n = 8,124,921), 

$-2\log(\Lambda)\sim \chi_{(J-1)(K-1)}^{2} \space approximately,$

$-2\log(\Lambda)=2\sum\limits_{j=1}^{J}\sum\limits_{k=1}^{K}y_{jk}\log\left(\frac{y_{jk}}{\hat{e}_{jk}}\right)$

$\hat{e}_{jk}=\frac{y_{j\bullet}y_{\bullet k}}{n}$

```{r G-Test}
results <- GTest(tab,correct="none")

print(results)
```

The p-value for this test is `r results$p.value` which is much less than 0.01. This suggests there is overwhelming evidence to reject the null hypothesis ($H_0$) and we can say the variables are not independent. This suggests that different areas have different levels of uptake of smart meters compared to others. 

